---
title: "Prediction Model for 30-Day Mortality After Acute Myocardial Infarction"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
theme: united
---

```{r setup, include=FALSE}
# This code runs when the document starts
# We only load the most essential packages
library(pROC)      # for ROC curves (simple to use)


```

# Introduction

Acute myocardial infarction (heart attack) is a serious condition. We want to predict which patients are at high risk of dying within 30 days. This can help doctors make better treatment decisions.

# Part 1: Initial Data Analysis

## a) Descriptives of the Key Variables

```{r load-data}
# Create our dataset with missing values
# In real life, you would read your data like this:
# gusto <- read.csv("gusto_data.csv")

set.seed(123)  # This makes our random numbers reproducible
n <- 1000      # We'll create 1000 patients

# Create the dataset with key variables only
gusto <- data.frame(
  # Outcome
  DAY30 = rbinom(n, 1, 0.1),           # 1 = died within 30 days, 0 = survived
  
  # Demographics
  AGE = round(runif(n, 19, 110)),      # Age in years
  SEX = rbinom(n, 1, 0.3),             # 1 = female, 0 = male
  
  # Clinical factors
  KILLIP = sample(1:4, n, replace = TRUE),  # Killip class (1-4, higher = worse)
  DIA = rbinom(n, 1, 0.15),             # 1 = diabetes, 0 = no diabetes
  HRT = rbinom(n, 1, 0.4),              # 1 = high heart rate, 0 = normal
  ANT = rbinom(n, 1, 0.4),              # 1 = anterior infarct, 0 = other location
  PMI = rbinom(n, 1, 0.2),              # 1 = previous heart attack, 0 = no
  
  # ECG findings
  STE = sample(0:11, n, replace = TRUE) # Number of leads with ST elevation (0-11)
)

# Add some missing values so we can practice handling them
gusto$AGE[sample(1:n, 50)] <- NA        # 50 missing ages
gusto$KILLIP[sample(1:n, 40)] <- NA     # 40 missing Killip class
gusto$DIA[sample(1:n, 35)] <- NA        # 35 missing diabetes
gusto$STE[sample(1:n, 45)] <- NA        # 45 missing ST elevation

# Look at the first few rows to understand our data
head(gusto)

# Get the structure of our data
str(gusto)

# Quick summary of all variables
summary(gusto)

# For categorical variables, we can use table()
table(gusto$SEX)
table(gusto$KILLIP)
table(gusto$DIA)
table(gusto$HRT)
table(gusto$ANT)
table(gusto$PMI)

# For continuous variables, we can calculate specific statistics
# AGE
mean(gusto$AGE, na.rm = TRUE)      # Mean age (ignore missing)
sd(gusto$AGE, na.rm = TRUE)        # Standard deviation
min(gusto$AGE, na.rm = TRUE)       # Minimum
max(gusto$AGE, na.rm = TRUE)       # Maximum
median(gusto$AGE, na.rm = TRUE)    # Median

# STE
mean(gusto$STE, na.rm = TRUE)
sd(gusto$STE, na.rm = TRUE)
min(gusto$STE, na.rm = TRUE)
max(gusto$STE, na.rm = TRUE)
median(gusto$STE, na.rm = TRUE)

# Let's create a simple summary table for continuous variables
# List our continuous variables
cont_vars <- c("AGE", "STE")

# METHOD 1: Using a for loop
summary_table_loop <- data.frame(
  Variable = cont_vars,
  Mean = NA,
  SD = NA,
  Median = NA,
  Min = NA,
  Max = NA,
  Missing = NA
)

for(i in 1:length(cont_vars)) {
  var_name <- cont_vars[i]
  var_data <- gusto[[var_name]]
  
  summary_table_loop$Mean[i] <- round(mean(var_data, na.rm = TRUE), 1)
  summary_table_loop$SD[i] <- round(sd(var_data, na.rm = TRUE), 1)
  summary_table_loop$Median[i] <- round(median(var_data, na.rm = TRUE), 1)
  summary_table_loop$Min[i] <- min(var_data, na.rm = TRUE)
  summary_table_loop$Max[i] <- max(var_data, na.rm = TRUE)
  summary_table_loop$Missing[i] <- sum(is.na(var_data))
}

summary_table_loop

# METHOD 2: Line by line (no loop)
age_data <- gusto$AGE
age_mean <- round(mean(age_data, na.rm = TRUE), 1)
age_sd <- round(sd(age_data, na.rm = TRUE), 1)
age_median <- round(median(age_data, na.rm = TRUE), 1)
age_min <- min(age_data, na.rm = TRUE)
age_max <- max(age_data, na.rm = TRUE)
age_missing <- sum(is.na(age_data))

ste_data <- gusto$STE
ste_mean <- round(mean(ste_data, na.rm = TRUE), 1)
ste_sd <- round(sd(ste_data, na.rm = TRUE), 1)
ste_median <- round(median(ste_data, na.rm = TRUE), 1)
ste_min <- min(ste_data, na.rm = TRUE)
ste_max <- max(ste_data, na.rm = TRUE)
ste_missing <- sum(is.na(ste_data))

summary_table_line <- data.frame(
  Variable = cont_vars,
  Mean = c(age_mean, ste_mean),
  SD = c(age_sd, ste_sd),
  Median = c(age_median, ste_median),
  Min = c(age_min, ste_min),
  Max = c(age_max, ste_max),
  Missing = c(age_missing, ste_missing)
)

summary_table_line

# For categorical variables, create frequency tables
cat_vars <- c("SEX", "KILLIP", "DIA", "HRT", "ANT", "PMI")

# Using a loop
for(var in cat_vars) {
  cat("\n---", var, "---\n")
  print(table(gusto[[var]], useNA = "ifany"))
  cat("Percentage:\n")
  print(round(prop.table(table(gusto[[var]], useNA = "ifany")) * 100, 1))
}

# Line by line example for first few
cat("\n--- SEX ---\n")
print(table(gusto$SEX, useNA = "ifany"))
print(round(prop.table(table(gusto$SEX, useNA = "ifany")) * 100, 1))

cat("\n--- KILLIP ---\n")
print(table(gusto$KILLIP, useNA = "ifany"))
print(round(prop.table(table(gusto$KILLIP, useNA = "ifany")) * 100, 1))
```

## b) Missing Values Analysis

```{r missing}
# Focus on key variables with missing data
key_vars <- c("AGE", "KILLIP", "DIA", "STE")

# METHOD 1: Using a loop
missing_table_loop <- data.frame(
  Variable = key_vars,
  Missing_Count = NA,
  Missing_Percent = NA
)

for(i in 1:length(key_vars)) {
  var_name <- key_vars[i]
  missing_count <- sum(is.na(gusto[[var_name]]))
  missing_pct <- round(mean(is.na(gusto[[var_name]])) * 100, 1)
  
  missing_table_loop$Missing_Count[i] <- missing_count
  missing_table_loop$Missing_Percent[i] <- missing_pct
}

missing_table_loop

# METHOD 2: Line by line
missing_AGE <- sum(is.na(gusto$AGE))
pct_AGE <- round(mean(is.na(gusto$AGE)) * 100, 1)

missing_KILLIP <- sum(is.na(gusto$KILLIP))
pct_KILLIP <- round(mean(is.na(gusto$KILLIP)) * 100, 1)

missing_DIA <- sum(is.na(gusto$DIA))
pct_DIA <- round(mean(is.na(gusto$DIA)) * 100, 1)

missing_STE <- sum(is.na(gusto$STE))
pct_STE <- round(mean(is.na(gusto$STE)) * 100, 1)

missing_table_line <- data.frame(
  Variable = key_vars,
  Missing_Count = c(missing_AGE, missing_KILLIP, missing_DIA, missing_STE),
  Missing_Percent = c(pct_AGE, pct_KILLIP, pct_DIA, pct_STE)
)

missing_table_line

# Check if missing data affects outcome
# Create missing indicators
gusto$AGE_missing <- is.na(gusto$AGE)
gusto$KILLIP_missing <- is.na(gusto$KILLIP)
gusto$DIA_missing <- is.na(gusto$DIA)
gusto$STE_missing <- is.na(gusto$STE)

# Compare mortality rates
# For AGE
rate_missing <- mean(gusto$DAY30[gusto$AGE_missing == TRUE], na.rm = TRUE)
rate_not_missing <- mean(gusto$DAY30[gusto$AGE_missing == FALSE], na.rm = TRUE)
cat("AGE - Mortality when missing:", round(rate_missing * 100, 1), "%\n")
cat("AGE - Mortality when not missing:", round(rate_not_missing * 100, 1), "%\n\n")

# For KILLIP
rate_missing <- mean(gusto$DAY30[gusto$KILLIP_missing == TRUE], na.rm = TRUE)
rate_not_missing <- mean(gusto$DAY30[gusto$KILLIP_missing == FALSE], na.rm = TRUE)
cat("KILLIP - Mortality when missing:", round(rate_missing * 100, 1), "%\n")
cat("KILLIP - Mortality when not missing:", round(rate_not_missing * 100, 1), "%\n\n")

# Using a loop for all missing variables
missing_vars <- c("AGE_missing", "KILLIP_missing", "DIA_missing", "STE_missing")
cat("\n=== Using loop ===\n")
for(var in missing_vars) {
  rate_missing <- mean(gusto$DAY30[gusto[[var]] == TRUE], na.rm = TRUE)
  rate_not <- mean(gusto$DAY30[gusto[[var]] == FALSE], na.rm = TRUE)
  cat(gsub("_missing", "", var), ":\n")
  cat("  Missing:", round(rate_missing * 100, 1), "%\n")
  cat("  Not missing:", round(rate_not * 100, 1), "%\n\n")
}
```

## c) Cross Tables: Outcome vs Key Predictors

```{r crosstabs}
# For categorical variables
cat_vars <- c("SEX", "KILLIP", "DIA", "HRT", "ANT", "PMI")

# METHOD 1: Using a loop
for(var in cat_vars) {
  cat("\n=== ", var, " ===\n")
  # Create table
  tbl <- table(gusto[[var]], gusto$DAY30, useNA = "ifany")
  print(tbl)
  
  # Row percentages
  cat("\nRow percentages (mortality rate by group):\n")
  pct <- round(prop.table(tbl, 1) * 100, 1)
  print(pct)
  
  # Chi-square test (to check if association is significant)
  if(nrow(tbl) > 1 && ncol(tbl) == 2 && sum(is.na(gusto[[var]])) == 0) {
    test <- chisq.test(tbl)
    cat("\nChi-square p-value:", round(test$p.value, 4), "\n")
  }
}

# METHOD 2: Line by line for key variables
cat("\n=== SEX ===\n")
tbl_sex <- table(gusto$SEX, gusto$DAY30, useNA = "ifany")
print(tbl_sex)
print(round(prop.table(tbl_sex, 1) * 100, 1))

cat("\n=== KILLIP ===\n")
tbl_killip <- table(gusto$KILLIP, gusto$DAY30, useNA = "ifany")
print(tbl_killip)
print(round(prop.table(tbl_killip, 1) * 100, 1))

cat("\n=== DIA (Diabetes) ===\n")
tbl_dia <- table(gusto$DIA, gusto$DAY30, useNA = "ifany")
print(tbl_dia)
print(round(prop.table(tbl_dia, 1) * 100, 1))

# For continuous variables, create groups
# AGE groups
gusto$AGE_group <- cut(gusto$AGE, 
                       breaks = c(0, 50, 65, 80, 120),
                       labels = c("<50", "50-65", "66-80", ">80"))

cat("\n=== AGE Groups ===\n")
tbl_age <- table(gusto$AGE_group, gusto$DAY30, useNA = "ifany")
print(tbl_age)
print(round(prop.table(tbl_age, 1) * 100, 1))

# STE groups (low, medium, high)
gusto$STE_group <- cut(gusto$STE,
                       breaks = c(-1, 2, 5, 12),
                       labels = c("0-2 leads", "3-5 leads", ">5 leads"))

cat("\n=== ST Elevation Groups ===\n")
tbl_ste <- table(gusto$STE_group, gusto$DAY30, useNA = "ifany")
print(tbl_ste)
print(round(prop.table(tbl_ste, 1) * 100, 1))
```

## d) Univariable Relationships (Single Predictor Models)

```{r univariable}
# Key predictors we want to test
key_predictors <- c("AGE", "SEX", "KILLIP", "DIA", "HRT", "ANT", "PMI", "STE")

# METHOD 1: Using a loop
# Create empty vectors to store results
or_values <- c()
ci_lower <- c()
ci_upper <- c()
p_values <- c()

for(i in 1:length(key_predictors)) {
  # Create formula
  formula <- as.formula(paste("DAY30 ~", key_predictors[i]))
  
  # Fit model
  model <- glm(formula, data = gusto, family = binomial)
  
  # Get results (if model converged)
  coef_sum <- summary(model)$coefficients
  
  if(nrow(coef_sum) > 1) {
    or_values[i] <- round(exp(coef_sum[2, 1]), 2)
    ci <- exp(confint(model)[2, ])
    ci_lower[i] <- round(ci[1], 2)
    ci_upper[i] <- round(ci[2], 2)
    p_values[i] <- round(coef_sum[2, 4], 4)
  } else {
    or_values[i] <- NA
    ci_lower[i] <- NA
    ci_upper[i] <- NA
    p_values[i] <- NA
  }
}

# Create results table
univariable_table <- data.frame(
  Predictor = key_predictors,
  Odds_Ratio = or_values,
  CI_lower = ci_lower,
  CI_upper = ci_upper,
  P_value = p_values
)

univariable_table

# METHOD 2: Line by line for each predictor

# AGE
model_age <- glm(DAY30 ~ AGE, data = gusto, family = binomial)
coef_age <- summary(model_age)$coefficients
or_age <- round(exp(coef_age[2, 1]), 2)
ci_age <- exp(confint(model_age)[2, ])
p_age <- round(coef_age[2, 4], 4)

# SEX
model_sex <- glm(DAY30 ~ SEX, data = gusto, family = binomial)
coef_sex <- summary(model_sex)$coefficients
or_sex <- round(exp(coef_sex[2, 1]), 2)
ci_sex <- exp(confint(model_sex)[2, ])
p_sex <- round(coef_sex[2, 4], 4)

# KILLIP
model_killip <- glm(DAY30 ~ KILLIP, data = gusto, family = binomial)
coef_killip <- summary(model_killip)$coefficients
or_killip <- round(exp(coef_killip[2, 1]), 2)
ci_killip <- exp(confint(model_killip)[2, ])
p_killip <- round(coef_killip[2, 4], 4)

# DIA (Diabetes)
model_dia <- glm(DAY30 ~ DIA, data = gusto, family = binomial)
coef_dia <- summary(model_dia)$coefficients
or_dia <- round(exp(coef_dia[2, 1]), 2)
ci_dia <- exp(confint(model_dia)[2, ])
p_dia <- round(coef_dia[2, 4], 4)

# Create table with first 4 predictors
univariable_line <- data.frame(
  Predictor = c("AGE", "SEX", "KILLIP", "DIA"),
  Odds_Ratio = c(or_age, or_sex, or_killip, or_dia),
  CI_lower = c(ci_age[1], ci_sex[1], ci_killip[1], ci_dia[1]),
  CI_upper = c(ci_age[2], ci_sex[2], ci_killip[2], ci_dia[2]),
  P_value = c(p_age, p_sex, p_killip, p_dia)
)

univariable_line

# Interpretation help
cat("\n=== INTERPRETATION ===\n")
cat("Odds Ratio > 1 means higher risk of death\n")
cat("Odds Ratio < 1 means lower risk of death\n")
cat("95% CI not crossing 1 means statistically significant\n")
cat("P-value < 0.05 means statistically significant\n")
```

# Part 2: Model Development

## a) Choosing Predictor Format

```{r choose_format}
# For this example, we'll use the continuous versions
# because they preserve more information

# We'll keep:
# - AGE (continuous) instead of creating age categories
# - KILLIP (continuous 1-4) because it's already ordered
# - STE (continuous 0-11) because number of leads matters

# Check if they are associated with outcome
cor(gusto$AGE, gusto$DAY30, use = "complete.obs")
cor(gusto$KILLIP, gusto$DAY30, use = "complete.obs")
cor(gusto$STE, gusto$DAY30, use = "complete.obs")
```

## b) Building a Multivariable Model

```{r building_model}
# Prepare data for modeling
gusto_model <- gusto

# Create missing indicators for key variables
gusto_model$AGE_miss <- ifelse(is.na(gusto_model$AGE), 1, 0)
gusto_model$KILLIP_miss <- ifelse(is.na(gusto_model$KILLIP), 1, 0)
gusto_model$DIA_miss <- ifelse(is.na(gusto_model$DIA), 1, 0)
gusto_model$STE_miss <- ifelse(is.na(gusto_model$STE), 1, 0)

# Simple imputation: replace missing with median (for continuous) or mode (for categorical)

# AGE (continuous) - impute with median
age_median <- median(gusto$AGE, na.rm = TRUE)
gusto_model$AGE[is.na(gusto_model$AGE)] <- age_median

# STE (continuous) - impute with median
ste_median <- median(gusto$STE, na.rm = TRUE)
gusto_model$STE[is.na(gusto_model$STE)] <- ste_median

# KILLIP (categorical) - impute with most common value
killip_table <- table(gusto$KILLIP)
killip_mode <- as.numeric(names(killip_table)[which.max(killip_table)])
gusto_model$KILLIP[is.na(gusto_model$KILLIP)] <- killip_mode

# DIA (binary) - impute with most common value
dia_table <- table(gusto$DIA)
dia_mode <- as.numeric(names(dia_table)[which.max(dia_table)])
gusto_model$DIA[is.na(gusto_model$DIA)] <- dia_mode

# HRT and others don't have missing in our example, but if they did:
# hrt_table <- table(gusto$HRT)
# hrt_mode <- as.numeric(names(hrt_table)[which.max(hrt_table)])
# gusto_model$HRT[is.na(gusto_model$HRT)] <- hrt_mode

# Now build our initial model with key predictors
# We'll include predictors that made clinical sense and had p < 0.2 in univariable analysis
initial_model <- glm(DAY30 ~ AGE + KILLIP + DIA + HRT + ANT + PMI + STE +
                     AGE_miss + KILLIP_miss + DIA_miss + STE_miss,
                     data = gusto_model, family = binomial)

# Look at the results
summary(initial_model)

# Simplify model by removing non-significant predictors (p > 0.05)
# Based on the summary, we might keep only significant ones
# This is a manual approach - we look at p-values and remove highest first

# For teaching, let's create a final model with the most important predictors
final_model <- glm(DAY30 ~ AGE + KILLIP + DIA + STE + AGE_miss,
                   data = gusto_model, family = binomial)

summary(final_model)
```

## c) Final Model Coefficients

```{r final_coef}
# Get detailed results from final model
coef_summary <- summary(final_model)$coefficients
conf_ints <- confint(final_model)

# Create a nice table
variables <- rownames(coef_summary)
results <- data.frame(
  Variable = variables,
  Coefficient = round(coef_summary[, 1], 3),
  `Std Error` = round(coef_summary[, 2], 3),
  `P-value` = round(coef_summary[, 4], 4),
  `Odds Ratio` = round(exp(coef_summary[, 1]), 2),
  `OR Lower` = round(exp(conf_ints[, 1]), 2),
  `OR Upper` = round(exp(conf_ints[, 2]), 2)
)

results

# Interpretation of each predictor
cat("\n=== INTERPRETATION OF RESULTS ===\n")
for(i in 2:nrow(results)) {
  cat("\n", results$Variable[i], ":\n")
  cat("  OR =", results$Odds.Ratio[i], 
      " (95% CI:", results$OR.Lower[i], "-", results$OR.Upper[i], ")\n")
  if(results$P.value[i] < 0.05) {
    cat("  This is statistically significant (p < 0.05)\n")
  } else {
    cat("  This is NOT statistically significant (p >= 0.05)\n")
  }
}
```

## d) Check for Important Interactions

```{r interactions}
# Test if age effect differs by diabetes status
model_int1 <- glm(DAY30 ~ AGE * DIA + KILLIP + STE + AGE_miss,
                  data = gusto_model, family = binomial)

# Compare with main effects model
anova(final_model, model_int1, test = "Chisq")

# Test if age effect differs by Killip class
model_int2 <- glm(DAY30 ~ AGE * KILLIP + DIA + STE + AGE_miss,
                  data = gusto_model, family = binomial)

anova(final_model, model_int2, test = "Chisq")

# If p-value > 0.05, interaction is not significant
```

## e) Regression Formula

```{r formula}
# Write out the model formula
coefs <- coef(final_model)

cat("Probability of death within 30 days = 1 / (1 + e^(-z))\n")
cat("where z =")

cat("\n", round(coefs[1], 3), " (intercept)")

for(i in 2:length(coefs)) {
  sign <- ifelse(coefs[i] > 0, " + ", " - ")
  cat(sign, abs(round(coefs[i], 3)), "×", names(coefs)[i])
}

# Example calculation for a specific patient
cat("\n\n=== EXAMPLE CALCULATION ===\n")
cat("Patient: AGE = 75, KILLIP = 3, DIA = 1, STE = 5, AGE_miss = 0\n")

z <- coefs[1] + 
     coefs[2] * 75 + 
     coefs[3] * 3 + 
     coefs[4] * 1 + 
     coefs[5] * 5 + 
     coefs[6] * 0

prob <- 1 / (1 + exp(-z))
cat("z =", round(z, 3), "\n")
cat("Predicted probability of death =", round(prob * 100, 1), "%\n")
```

## f) Simple Clinical Risk Score

```{r clinical}
# Create a simplified scoring system
# Multiply coefficients by 10 and round to make integers
coefs <- coef(final_model)[-1]  # exclude intercept

scores <- round(coefs * 10, 0)

score_chart <- data.frame(
  Predictor = names(coefs),
  Points = scores,
  How_to_use = c("Per 10 years", "Per 1 class increase", "1 if yes, 0 if no", 
                 "Per lead", "1 if age missing")
)

score_chart

# Calculate risk score for each patient
gusto_model$risk_score <- 
  scores[1] * (gusto_model$AGE / 10) +  # Age per 10 years
  scores[2] * gusto_model$KILLIP +
  scores[3] * gusto_model$DIA +
  scores[4] * gusto_model$STE +
  scores[5] * gusto_model$AGE_miss

# Look at relationship between score and actual outcome
# Create score groups
gusto_model$risk_group <- cut(gusto_model$risk_score, 
                              breaks = quantile(gusto_model$risk_score, 
                                               probs = seq(0, 1, 0.2)),
                              labels = c("Very Low", "Low", "Medium", "High", "Very High"))

# Calculate mortality in each group
for(group in levels(gusto_model$risk_group)) {
  idx <- gusto_model$risk_group == group & !is.na(gusto_model$risk_group)
  mortality <- mean(gusto_model$DAY30[idx], na.rm = TRUE)
  n_patients <- sum(idx)
  cat(group, ": n =", n_patients, ", mortality =", round(mortality * 100, 1), "%\n")
}
```

# Part 3: Validation and Performance

## a) Model Performance

```{r performance}
# Calculate predicted probabilities
gusto_model$pred_prob <- predict(final_model, type = "response")

# ROC curve
roc_curve <- roc(gusto_model$DAY30, gusto_model$pred_prob)
plot(roc_curve, main = "ROC Curve for 30-Day Mortality Prediction")
text(0.8, 0.2, paste("AUC =", round(auc(roc_curve), 3)))

auc_value <- auc(roc_curve)
auc_ci <- ci.auc(roc_curve)

cat("=== DISCRIMINATION (How well model separates survivors from non-survivors) ===\n")
cat("AUC:", round(auc_value, 3), "\n")
cat("95% CI:", round(auc_ci[1], 3), "-", round(auc_ci[3], 3), "\n")
cat("Interpretation:\n")
cat("  AUC = 0.5: No better than chance\n")
cat("  AUC = 0.7-0.8: Acceptable\n")
cat("  AUC = 0.8-0.9: Excellent\n")
cat("  AUC > 0.9: Outstanding\n\n")

# Calibration: how well predicted probabilities match observed outcomes
# Create 10 groups based on predicted risk
gusto_model$risk_group <- cut(gusto_model$pred_prob, 
                              breaks = quantile(gusto_model$pred_prob, 
                                               probs = seq(0, 1, 0.1)),
                              include.lowest = TRUE)

# Calculate mean predicted and observed for each group
calibration_table <- data.frame(
  Group = 1:10,
  N = NA,
  Predicted = NA,
  Observed = NA
)

for(i in 1:10) {
  # This is a bit complex, but it gets the groups correctly
  group_name <- levels(gusto_model$risk_group)[i]
  idx <- gusto_model$risk_group == group_name & !is.na(gusto_model$risk_group)
  
  calibration_table$N[i] <- sum(idx)
  calibration_table$Predicted[i] <- round(mean(gusto_model$pred_prob[idx], na.rm = TRUE), 3)
  calibration_table$Observed[i] <- round(mean(gusto_model$DAY30[idx], na.rm = TRUE), 3)
}

cat("=== CALIBRATION (How well predicted risks match actual outcomes) ===\n")
print(calibration_table)

# Simple calibration plot
plot(calibration_table$Predicted, calibration_table$Observed, 
     xlab = "Predicted Risk", ylab = "Observed Risk",
     main = "Calibration Plot", 
     xlim = c(0, max(calibration_table$Predicted, calibration_table$Observed)),
     ylim = c(0, max(calibration_table$Predicted, calibration_table$Observed)),
     pch = 19, col = "blue")
abline(0, 1, col = "red", lwd = 2)  # Perfect calibration line
legend("topleft", legend = "Perfect calibration", col = "red", lwd = 2)
```

## b) Simple Validation (Split-sample)

```{r validation}
# Split data into training (70%) and validation (30%)
set.seed(123)
n_train <- round(0.7 * nrow(gusto_model))
train_idx <- sample(1:nrow(gusto_model), n_train, replace = FALSE)

train_data <- gusto_model[train_idx, ]
valid_data <- gusto_model[-train_idx, ]

# Refit model on training data
train_model <- glm(DAY30 ~ AGE + KILLIP + DIA + STE + AGE_miss,
                   data = train_data, family = binomial)

# Apply to validation data
valid_data$pred_prob <- predict(train_model, newdata = valid_data, type = "response")

# Calculate AUC in validation set
valid_roc <- roc(valid_data$DAY30, valid_data$pred_prob)
valid_auc <- auc(valid_roc)

cat("=== VALIDATION RESULTS ===\n")
cat("Training set AUC:", round(auc(roc(train_data$DAY30, 
                                        predict(train_model, type = "response"))), 3), "\n")
cat("Validation set AUC:", round(valid_auc, 3), "\n")
cat("Difference:", round(abs(valid_auc - auc_value), 3), "\n")
cat("Small difference (<0.05) suggests model is not overfitted\n")
```

# Part 4: Conclusions

```{r conclusions}
cat("=== SUMMARY OF FINDINGS ===\n\n")

cat("1. DATA DESCRIPTION:\n")
cat("   - Total patients:", nrow(gusto), "\n")
cat("   - Mortality rate:", round(mean(gusto$DAY30, na.rm = TRUE) * 100, 1), "%\n")
cat("   - Key predictors: Age, Killip class, Diabetes, ST elevation\n\n")

cat("2. MODEL PERFORMANCE:\n")
cat("   - AUC:", round(auc_value, 3), "\n")
cat("   - This means the model can distinguish between\n")
cat("     survivors and non-survivors", 
    ifelse(auc_value > 0.7, "reasonably well", "poorly"), "\n\n")

cat("3. KEY PREDICTORS:\n")
for(i in 2:nrow(results)) {
  cat("   -", results$Variable[i], ": ")
  cat("OR =", results$Odds.Ratio[i])
  if(results$P.value[i] < 0.05) cat(" (significant)")
  cat("\n")
}

cat("\n4. CLINICAL USE:\n")
cat("   - The model can help identify high-risk patients\n")
cat("   - Simple scoring system:\n")
for(i in 1:nrow(score_chart)) {
  cat("     * ", score_chart$Predictor[i], ": ", score_chart$Points[i], 
      " points ", score_chart$How_to_use[i], "\n", sep="")
}

cat("\n5. LIMITATIONS:\n")
cat("   - Missing data required imputation\n")
cat("   - Need validation in other populations\n")
cat("   - Model should be updated with new data\n")

cat("\n6. NEXT STEPS:\n")
cat("   - Test in other hospitals\n")
cat("   - Create a simple calculator app for bedside use\n")
cat("   - Consider adding other important predictors\n")
cat("   - Update model as new data becomes available\n")
```

# Appendix: R Functions Used in This Analysis

```{r appendix, eval=FALSE}
# BASIC FUNCTIONS:
# head()        - View first few rows
# str()         - Check data structure
# summary()     - Quick statistics
# table()       - Frequency tables
# prop.table()  - Convert to percentages
# mean()        - Calculate average
# sd()          - Standard deviation
# min(), max()  - Minimum and maximum
# median()      - Median value
# quantile()    - Percentiles
# is.na()       - Check for missing values
# colSums()     - Sum of columns

# DATA MANIPULATION:
# data.frame()  - Create data frame
# names()       - Get column names
# rownames()    - Get row names
# [ , ]         - Select rows and columns
# $             - Select a column
# ifelse()      - Conditional statements
# cut()         - Create categories from continuous

# STATISTICS:
# glm()         - Logistic regression
# summary()     - Model results
# coef()        - Extract coefficients
# confint()     - Confidence intervals
# exp()         - Exponential (for odds ratios)
# predict()     - Make predictions
# roc()         - Create ROC curve (from pROC)
# auc()         - Get AUC value
# ci.auc()      - Confidence interval for AUC
# anova()       - Compare models
# chisq.test()  - Chi-square test

# PROGRAMMING:
# for()         - Loops
# paste()       - Combine text
# as.formula()  - Create formula from text
# c()           - Combine values
# length()      - Length of vector
# nrow(), ncol() - Get dimensions
```
